The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-18 paper.tex
(*) review

\begin{itemize}
\item Relevancia de Esterel
\item Ceu
    \begin{itemize}
        \item Foco inicial em "constrained WSNs"
        \item Movendo para outros dominios? (jogos, artigo Mod'15)
        \item Usos: SenSys, Terra, sala de aula, GSoC
    \end{itemize}
\item Limitacoes do modelo sincrono
\item Semantica e Implementacao (nesse artigo, somente subset estatico)
\end{itemize}

@@

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In the synchronous model, programs advance in a sequence of discrete reactions 
to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.

@@

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{"call 
f1() || call f2()"}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
In \CEU, when multiple trails are active at a time, they are scheduled in the 
order they appear in the program source code.
%
Hence, \CEU is deterministic also with respect to the order of execution of 
side effects within a reaction.
%
Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel and \CEU to illustrate the semantic difference regarding 
(non-)determinism.

On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
In contrast, Esterel precludes support for shared memory: \emph{``if a variable 
is written by some thread, then it can neither be read nor be written by 
concurrent threads''}~\cite{esterel.primer}.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other embedded languages made a similar design choice~\cite{wsn.sol,pret}.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

@@

\begin{figure}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SENDACK;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                _send_enqueue(&buffer);
                await SENDACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.41\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        _send_enqueue(&buffer);
    with
        _send_cancel(&buffer);
    end
    await SENDACK;
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{A network protocol in \CEU extended with a finalization clause to 
cancel a transmission.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

The example in the left of Figure~\ref{lst.fin} relies on hierarchical 
\code{par/or} and \code{par/and} compositions to describe the state machine of 
a data collection protocol for sensor networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol with a client application.
%
The protocol has two trails:
one monitors the stopping event (line 3);
the other periodically transmits a packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles an immediate retransmission request (line 7);
the other transmits the packet awaiting for a confirmation (lines 9--16).
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, to avoid flooding the network with 
packets.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Note that the \code{buffer} packet (line 12) is a local variable whose address 
is passed to function \code{\_send\_enqueue} for transmission.
The call enqueues the pointer in the radio driver, which holds it up to the 
emission of \code{SENDACK} acknowledging the packet transmission.
%
In the meantime, if the sending trail is be aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

%Programs in \CEU can access \emph{C} libraries available in the underlying 
%platform by prefixing symbols with an underscore (e.g., 
%\code{\_printf(<...>)}).
%
The \CEU compiler tracks the interaction of \code{par/or} compositions with 
local variables and stateful \emph{C} functions (e.g., device drivers) in order 
to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
For instance, \CEU refuses to compile the program in the left of 
Figure~\ref{lst.fin}, enforcing the programmer to write a \emph{finalization} 
clause to accompany the stateful \emph{C} call~\cite{ceu.sensys13}.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \emph{C} libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

Activities that involve reactions to \emph{wall-clock time}%
\footnote{
By wall-clock time we mean the passage of time from the real world, measured in 
hours, minutes, etc.
}
appear in typical patterns of embedded development, such as timeouts and sensor 
sampling.
However, support for wall-clock time is somewhat low-level in existing 
languages, usually through timer callbacks or ``sleep'' blocking calls.
%
However, in any concrete system implementation, a requested timeout does not 
expire precisely with zero-delay, a fact that is usually ignored in the 
development process.
We define the difference between the requested timeout and the actual expiring 
time as the \emph{residual delta time (delta)}.
Without explicit manipulation, the recurrent use of timed activities in 
sequence (or in a loop) may accumulate a considerable amount of deltas that can 
lead to incorrect behavior in programs.

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
For the example in the left of Figure~\ref{lst.timers}, suppose that after the 
first \code{await} request, the underlying system gets busy and takes 15ms to 
check for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during the entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, however, internal events follow a stack-based execution policy, 
similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
and they are both become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st]}
\item 2nd trail awakes (line 9), calls \code{\_f(1)}, and terminates.\\
    \emph{stack: [1st]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f(2)}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

\begin{figure}[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

@@

On the one hand, this form of subroutines has a significant limitation that it 
cannot express recursive calls: an \code{emit} to itself is always ignored, 
given that a running body cannot be awaiting itself.
%
On the other hand, this very same limitation brings some important safety 
properties to subroutines:
first, they are guaranteed to react in bounded time;
second, memory for locals is also bounded, not requiring data stacks.
%
Also, this form of subroutines can use the other primitives of \CEU, such as 
parallel compositions and the \code{await} statement.
In particular, they await keeping context information such as locals and the 
program counter, just like coroutines~\cite{lua.coroutines}.
%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
%exceptions.

@@

%The disciplined synchronous execution of \CEU, together with broadcast 
%communication and stacked execution for internal events, may raise doubts 
%%about the precise execution of programs.
%
In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics in order to formally describe the language.
%, eliminating imprecisions with regard to how a program reacts to an external 
%event.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and $C$ calls (which behave like in any conventional 
imperative language).

%The full syntax and semantic rules can be obtained in 
%Appendix~\ref{app.formal}.

%In section~\ref{sec.formal.ceu}, we show how to translate \CEU statements into

@@

\begin{figure}[h]
%\rule{8.5cm}{0.37pt}
{\small
\begin{verbatim}
                                   // primary expressions
  p ::= mem(id)                    (any memory access to `id')
      | await(id)                  (await event `id')
      | emit(id)                   (emit event `id')
      | break                      (loop escape)
                                   // compound expressions
      | if mem(id) then p else p   (conditional)
      | p ; p                      (sequence)
      | loop p                     (repetition)
      | p and p                    (par/and)
      | p or p                     (par/or)
      | fin p                      (finalization)
                                   // derived by semantic rules
      | awaiting(id,n)             (awaiting `id' since sequence number `n')
      | emitting(n)                (emitting on stack level `n')
      | p @ loop p                 (unwinded loop)
\end{verbatim}
}%
\caption{
    Reduced syntax of \CEU.
\label{lst.formal.syntax}
}
\end{figure}

Figure~\ref{lst.formal.syntax} shows the BNF-like syntax for a subset of \CEU 
that is sufficient to describe all semantic peculiarities of the language.
%
The $mem(id)$ primitive represents all accesses, assignments, and $C$ function 
calls that affect a memory location identified by $id$.
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
%considered to be atomic and instantaneous
%
The special notation $nop$ is used to represent an innocuous $mem$ expression 
(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
unused identifier).
%
Except for the $fin$ and semantic-derived expressions, which are discussed 
further, the other expressions map to their counterparts in the concrete 
language in Figure~\ref{lst.syntax}.
%
Note that $mem$ expressions cannot share identifiers with $await$/$emit$ 
expressions.

@@

The core of our semantics is a relation that, given a sequence number $n$ 
identifying the current reaction chain, maps a program $p$ and a stack of 
events $S$ in a single step to a modified program and stack:
%
$$
\LL S, p \RR
    \xrightarrow[~~n~~]{}
\LL S', p' \RR
$$
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(sequence~of~event~identifiers: [id_{top}, ..., id_1]) \\
p, p' &\in P
    && (as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (univocally~identifies~a~reaction~chain)
\end{align*}

At the beginning of a reaction chain, the stack is initialized with the 
occurring external event $ext$ ($S=[ext]$), but $emit$ expressions can push new 
events on top of it (we discuss how they are popped further).

@@

We describe this relation with a set of \emph{small-step} structural semantics 
rules, which are built in such a way that at most one transition is possible at 
any time, resulting in deterministic reaction chains.
%
The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n) \RR
    & \textbf{(await)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2m<n
    & \textbf{(awaiting)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emit)}       \\
%%%
\LL S,\1emitting(|S|) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(emitting)}
\end{align*}
}

An $await$ is simply transformed into an $awaiting$ that remembers the current 
external sequence number $n$ (rule \textbf{await}).
An $awaiting$ can only transit to a $nop$ (rule \textbf{awaiting}) if its 
referred event $id$ matches the top of the stack and its sequence number is 
smaller than the current one ($m<n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (in 
rule \textbf{emit}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{emitting} expects its parameter to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

@@

Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives the memory identifier and current 
reaction sequence number, returning the current memory value.
%
Although the value is arbitrary, it is unique in a reaction chain, because a 
given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

When a program first encounters a $loop$, it first expands its body in sequence 
with itself (rule \textbf{loop-expd}).
Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
\textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
$mem(id)$.
However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
Note that if we used \code{`;'} as a separator in loops, rules 
\textbf{loop-brk} and \textbf{seq-brk} would conflict.
%
Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
a $nop$.
%Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
%into a $clear(p)$.
%We cannot simply transform the loop into a $nop$ because its body may be a 
%parallel composition containing finalization blocks.

The rules for parallel $and$ compositions force transitions on the left branch 
$p$ to occur before transitions on the right branch $q$ (rules 
\textbf{and-adv1} and \textbf{and-adv2}).
Then, if one of the sides terminates, the composition is simply substituted by 
the other side (rules \textbf{and-nop1} and \textbf{and-nop2}):

>>>>

\begin{itemize}
\item Relevancia de Esterel
\item Ceu
    \begin{itemize}
        \item Foco inicial em "constrained WSNs"
        \item Estendendo para outros dominios? (jogos, artigo Mod'15)
        \item Usos: SenSys, Terra, sala de aula, GSoC
    \end{itemize}
\item Limitacoes do modelo sincrono
\item Semantica e Implementacao (nesse artigo, somente subset estatico)
% Artigo de Rust, nao resolve os problemas
\end{itemize}

@@

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in whicn programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.

@@

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{"call 
f1() || call f2()"}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
In \CEU, when multiple trails are active at a time, they are scheduled in the 
order they appear in the program source code.
%
Hence, \CEU is deterministic also with respect to the order of execution of 
side effects within a reaction~\cite{ceu.sensys13}.
%
Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel and \CEU to illustrate the semantic difference regarding 
(non-)determinism.

On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
In contrast, Esterel forbids sharing memory: \emph{``if a variable is written 
by some thread, then it can neither be read nor be written by concurrent 
threads''}~\cite{esterel.primer}.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other embedded languages made a similar design choice~\cite{wsn.sol,pret}.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

@@

\begin{figure}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SENDACK;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                _send_enqueue(&buffer);
                await SENDACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.41\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        _send_enqueue(&buffer);
    with
        _send_cancel(&buffer);
    end
    await SENDACK;
<...>
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{A network protocol in \CEU extended with a finalization clause to 
cancel a transmission.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

To illustrate threats related to abortion, consider the example in the left of 
Figure~\ref{lst.fin} with hierarchical \code{par/or} and \code{par/and} 
compositions describing the state machine of a data collection protocol for 
sensor networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol with a client application.
%
The protocol has two trails:
one simply monitors the stopping event (line 3);
the other periodically transmits a packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles an immediate retransmission request (line 7);
the other transmits the packet%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \emph{C} libraries explicit and are required in \CEU.
}
and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, to avoid flooding the network with 
packets.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Note that the \code{buffer} packet (line 12) is a local variable whose address 
is passed to function \code{\_send\_enqueue} for transmission.
The call enqueues the pointer in the radio driver, which holds it up to the 
emission of \code{SENDACK} acknowledging the packet transmission.
%
In the meantime, if the sending trail is be aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The \CEU compiler tracks the interaction of \code{par/or} compositions with 
local variables and stateful \emph{C} functions (e.g., device drivers) in order 
to preserve safe abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
For instance, \CEU refuses to compile the program in the left of 
Figure~\ref{lst.fin}, enforcing the programmer to write a \emph{finalization} 
clause to accompany the stateful \emph{C} call.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \emph{C} libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

Activities that involve reactions to \emph{wall-clock time}%
\footnote{
By wall-clock time we mean the passage of time from the real world, measured in 
hours, minutes, etc.
}
appear in typical patterns of embedded development, such as timeout watchdogs 
and sensor samplings.
However, support for wall-clock time is somewhat low-level in existing 
languages, usually through timer callbacks or ``sleep'' blocking calls.
%
Furthermore, in any concrete timer implementation, a requested timeout does not 
expire precisely without delays, a fact that is usually ignored in the 
development process.
We define the difference between the requested timeout and the actual expiring 
time as the \emph{residual delta time (delta)}.
Without explicit manipulation, the recurrent use of timed activities in 
sequence (or in a loop) may accumulate a considerable amount of deltas that can 
lead to incorrect behavior in programs.

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically~\cite{ceu.sensys13}, resulting in more robust applications.
For the example in the left of Figure~\ref{lst.timers}, suppose that after the 
first \code{await} request, the underlying system gets busy and takes 15ms to 
check for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during the entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, however, internal events follow a stack-based execution policy, 
similar to subroutine calls in typical programming 
languages~\cite{ceu.sensys13}, contrasting with queue-based event handling 
adopted for external events.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
and they both become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st]}
\item 2nd trail awakes (line 9), calls \code{\_f()}, and terminates.\\
    \emph{stack: [1st]}
\item 1st trail (on top of the stack) resumes, calls \code{\_g()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

\begin{figure}%[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

@@

On the one hand, this form of subroutines has a significant limitation that it 
cannot express recursive calls: an \code{emit} to itself is always ignored, 
given that a running body cannot be awaiting itself.
%
On the other hand, this very same limitation brings some important safety 
properties to subroutines:
first, they are guaranteed to react in bounded time;
second, memory for locals is also bounded, not requiring data stacks.
%
Also, this form of subroutines can use the other primitives of \CEU, such as 
parallel compositions and the \code{await} statement.
In particular, they await keeping context information such as locals and the 
program counter, similarly to coroutines~\cite{lua.coroutines}.
%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
%exceptions.

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the main differences to Esterel, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and $C$ calls (which behave like in any conventional 
imperative language).

@@

\begin{figure}[h]
\begin{lstlisting}[mathescape=true]
                                   // primary expressions
  p ::= mem(id)                   (any memory access to `id')
      $|$ await(id)                  (await event `id')
      $|$ emit(id)                   (emit event `id')
      $|$ break                      (loop escape)
                                   // compound expressions
      $|$ if mem(id) then p else p   (conditional)
      $|$ p ; p                      (sequence)
      $|$ loop p                     (repetition)
      $|$ p and p                    (par/and)
      $|$ p or p                     (par/or)
      $|$ fin p                      (finalization)
                                   // derived by semantic rules
      $|$ awaiting(id,n)             (awaiting `id' since sequence number `n')
      $|$ emitting(n)                (emitting on stack level `n')
      $|$ p @ loop p                 (unwinded loop)
\end{lstlisting}
\caption{
    Reduced syntax of \CEU.
\label{lst.formal.syntax}
}
\end{figure}

Figure~\ref{lst.formal.syntax} shows the BNF-like syntax for a subset of \CEU 
that is sufficient to describe all semantic peculiarities of the language.
%
Except for $fin$ and all semantic-derived expressions (i.e., $awating$, 
$emitting$, and $p~@~loop; p$), which are discussed further, all other 
expressions map to their counterparts in the concrete language.

The $mem(id)$ primitive represents all accesses, assignments, and $C$ function 
calls that affect a memory location identified by $id$.
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
%considered to be atomic and instantaneous
%
The special notation $nop$ is used to represent an innocuous $mem$ expression 
(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

@@

The core of our semantics is a relation that, given a sequence number $n$ 
identifying the current reaction chain, maps a program $p$ and a stack of 
events $S$ in a single step to a modified program and stack:
%
\begin{align*}
\LL S, p \RR &\ST
\LL S', p' \RR
    & \textbf{(relation-inner)}
\end{align*}
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(sequence~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
p, p' &\in P
    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (unique~identifier~for~the~reaction~chain)
\end{align*}

At the beginning of a reaction chain, the stack is initialized with the 
occurring external event $ext$ ($S=[ext]$), but $emit$ expressions can push new 
events on top of it (we discuss how they are popped further).
The ever-increasing sequence number $n$ will prevent that awaiting expressions 
awake in the same reaction they are reached (ensuring that the program can 
progress).

@@

We describe this relation with a set of \emph{small-step} structural semantics 
rules, which are built in such a way that at most one transition is possible at 
any time, resulting in deterministic reaction chains.
%
The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n+1) \RR
    & \textbf{(awaiting)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2m<=n
    & \textbf{(awake)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emitting)}       \\
%%%
\LL S,\1emitting(|S|) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(pop)}
\end{align*}
}

An $await$ is simply transformed into an $awaiting$ (rule \textbf{awaiting}) as 
an artifact to remember the external sequence number $n$ it can awake.
An $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
event $id$ matches the top of the stack and its sequence number is smaller or 
equal than the current one ($m<n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (rule 
\textbf{emitting}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{pop} expects its parameter to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

@@

Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives the memory identifier and current 
reaction sequence number, returning the current memory value.
%
Although the value here is arbitrary, it is unique in a reaction chain, because 
a given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{awaiting}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

When a program encounters a $loop$, it first expands its body in sequence with 
itself (rule \textbf{loop-expd}).
Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
\textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
$mem(id)$.
However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
Note that if we used \code{`;'} as a separator in loops, rules 
\textbf{loop-brk} and \textbf{seq-brk} would conflict.
%
Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
a $nop$.
%Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
%into a $clear(p)$.
%We cannot simply transform the loop into a $nop$ because its body may be a 
%parallel composition containing finalization blocks.

Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
always force transitions on their left branches $p$ to occur before their right 
branches $q$:

{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
    & \textbf{(or-adv2)}   \\
\end{eqnarray*}
}
